European or American? A re-examination of the transatlantic influence over the IMF

### 2025-07-25 manuscript/manuscript.qmd
Add WB comparison
Add original section on IMF economic lendnig criteria in red

For the strand of literature that examines American and European influence separately, nearly all studies identify the United States as the dominant actor within the IMF [@stone2011, p.51]. Qualitative accounts emphasise the United States’ foundational role in shaping the IMF’s governance structure [@stone2011], the Fund’s practice of making direct enquiries with Washington for major lending decisions [@southard1979; @woods2003], and its unique status as the only country-level veto player, stemming from its position as the largest financial contributor [@woods2003; @stone2011]. The US also benefits from significant informational advantages over other members [@stone2011, p.57] and proximity-based access, with the IMF headquartered in Washington, D.C. [@stone2011, p.57]. Numerous case studies further demonstrate instances of US political pressure resulting in lenient loan conditions for key geopolitical allies, including Zaire, Egypt, Mexico, Argentina, and South Korea [@vreeland2009; @copelovitch2010a; @lipscy2018].

@@

The projection of great power capabilities into international organizations is sometimes uneven, because of strategic preferences, the perceived inefficiency of institutional engagement, or other possible reasons. Existing studies provide substantial evidence of U.S. influence over IMF lending decisions, they leave a crucial question open-the comparison between the U.S. and other major powers, particularly the European countries. IMF is perceived to be an institution dominated by a few major power, but the relative weight of American versus European remains ambiguous. The organization is traditionally chaired by an European national [@kahler2001],

>>>>

For the strand of literature that examines American and European influence separately, nearly all studies identify the United States as the dominant actor within the IMF [@stone2011, p.51]. Qualitative accounts emphasise the United States’ foundational role in shaping the IMF’s governance structure [@stone2011], the Fund’s practice of making direct enquiries with Washington for major lending decisions [@southard1979; @woods2003], and its unique status as the only country-level veto player, stemming from its position as the largest financial contributor [@woods2003; @stone2011]. The US also benefits from significant informational advantages over other members [@stone2011, p.57] and proximity-based access, with the IMF headquartered in Washington, D.C. [@stone2011, p.57]. Numerous case studies further demonstrate instances of US political pressure resulting in lenient loan conditions for key geopolitical allies, including Zaire, Egypt, Mexico, Argentina, and South Korea [@vreeland2009; @copelovitch2010a; @lipscy2018]. This pattern of influence is consistent with the United States’ broader role across international financial institutions, particularly the World Bank, where American dominance is widely acknowledged [@wade2002; @woods2003; @clark2020].

@@

\color{red}
However, empirical studies suggested that countries with low international reserves, high current account deficits, and low GDP growth are more likely to receive IMF funding assistance [@barro2005; @knight1997], despite such indicators often reflecting a limited capacity to repay.
\color{black}
